An open and continuously updated fern tree of life (FTOL)

### 2022-02-15 ms/manuscript.Rmd
Move calculation of coverage to targets pipeline

# Targets from MS workflow
tar_load(
	c(
		sanger_sampling,
		gb_species_by_year,
		ref_files,
		other_dates,
		crown_ages,
		family_stem_ages
	),
	store = here::here("_targets")
)
```

@@

```{r sampling-stats, message=FALSE}
# FTOL sampling ----

# Tally FTOL sampling by rank
ftol_sampling_by_rank <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	# drop suborder
	select(-suborder) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup

# Make list of all accepted species
accepted_species <-
	pteridocat %>% 
	# Only keep accepted names
	filter(str_detect(taxonomicStatus, "accepted")) %>%
	# Only keep species
	# FIXME: need to add rank for newly added names in pteridocat_maker
	# then don't need is.na()
	filter(taxonRank == "species" | is.na(taxonRank)) %>%
	select(scientificName) %>%
	mutate(
		rgnparser::gn_parse_tidy(scientificName) %>% 
			select(species = canonicalsimple),
		species = str_replace_all(species, " ", "_")
	) %>%
	# Map on higher level taxonomy
	mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
	left_join(
		select(ppgi_taxonomy, genus, family, 
					 suborder, order, class), by = "genus"
	) %>%
	# FIXME: update class for nothogenera in PPG
	mutate(class = case_when(
		str_detect(
			scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
		TRUE ~ class
	)) %>%
	# Drop lycophytes
	assert(not_na, class) %>%
	filter(class == "Polypodiopsida") %>%
	select(-class) %>%
	# Add major_clade
	add_major_clade() %>%
	# drop suborder
	select(-suborder)

# Tally accepted sampling by rank
accepted_sampling_by_rank <-
	accepted_species %>%
	select(-scientificName) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup %>%
	# Note that nothogenera have no family or order
	filter(!is.na(name))

# Format coverage at order-ish level for plotting
coverage_data <- left_join(
	ftol_sampling_by_rank,
	rename(accepted_sampling_by_rank, n_accepted = n),
	by = c("rank", "name")
) %>%
	mutate(coverage = n/n_accepted)

# Format for text in MS
tax_coverage <-
	coverage_data %>%
	filter(rank == "major_clade") %>%
	arrange(coverage) %>%
	transmute(clade = name, coverage = percent(coverage, accuracy = 0.1)) %>%
	split(.$clade)

# Format overall sampling for printing in MS
total_sampling <-
	# Start with FTOL sampling by genus, family, order
	ftol_sampling_by_rank %>% 
	count(rank) %>%
	# Add n species
	bind_rows(
		sanger_sampling %>%
			filter(outgroup == FALSE) %>%
			select(-outgroup) %>%
			summarize(n = n(), rank = "species")
	) %>%
	# Join to accepted number of species, genus, family, order
	left_join(
		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
			bind_rows(
				summarize(accepted_species, n_accepted = n(), rank = "species")
			),
		by = "rank"
	) %>%
	# Calculate coverage
	mutate(
		coverage_p = percent(n / n_accepted, accuracy = 0.1),
		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")) %>% #nolint
	# Split for printing
	split(.$rank)

# Plastome (backbone) sampling ----
bb_sampling <-
	tibble(species = plastome_tree$tip.label) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species")

bb_sampling_by_rank <-
	bb_sampling %>%
	# Drop OG
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup

# Format sampling for printing in MS
bb_total_sampling <-
	# Start with backbone sampling by genus, family, order
	bb_sampling_by_rank %>% 
	count(rank) %>%
	# Add n species
	bind_rows(
		bb_sampling %>%
			filter(outgroup == FALSE) %>%
			select(-outgroup) %>%
			summarize(n = n(), rank = "species")
	) %>%
	# Join to accepted number of species, genus, family, order
	left_join(
		count(accepted_sampling_by_rank, rank, name = "n_accepted") %>%
			bind_rows(
				summarize(accepted_species, n_accepted = n(), rank = "species")
			),
		by = "rank"
	) %>%
	# Calculate coverage
	mutate(
		coverage_p = percent(n / n_accepted),
		coverage = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 0.1)}")) %>% #nolint
	# Split for printing
	split(.$rank)
```

@@

plastome_gleich_genera_n <-
	bb_sampling %>%
	filter(major_clade == "Gleicheniales") %>%
	pull(genus) %>%
	n_distinct()
```

@@

clade_cols_tib <-
	tibble(
		color = clade_cols,
		major_clade = names(clade_cols)
	) %>%
	mutate(
		code = substr(major_clade, 1, 2),
		code = case_when(
			major_clade == "Saccolomatineae" ~ "Sac",
			major_clade == "Salviniales" ~ "Sal",
			TRUE ~ code
		)) %>%
	assert(is_uniq, code)

```

@@

# Format coverage at order-ish level for plotting
b <-
	coverage_data %>%
	filter(rank == "major_clade") %>%
	left_join(major_clade_exemp_taxa, by = c(name = "major_clade")) %>%
	arrange(phy_order) %>%
	rename(clade = name) %>%
	mutate(
		clade = fct_reorder(clade, phy_order) %>%
			fct_rev(),
		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
	) %>%
	ggplot(aes(x = coverage, y = clade, fill = clade)) +
	geom_col(color = "grey10", size = 0.2) +
	geom_text(
		aes(label = coverage_text),
		hjust = 0,
		nudge_x = 0.01,
		size = 6*.36 # 6 pt font
	) +
	scale_x_continuous(
		labels = percent,
		limits = c(0,1),
		expand = expansion(mult = c(0, 0))
	) +
	scale_fill_manual(values = clade_cols) +
	scale_y_discrete(labels = clade_labs) +
	labs(x = "Coverage") +
	theme_bw() +
	theme(
		axis.text.y = element_markdown(size = 6),
		axis.title.y = element_blank(),
		legend.position = "none",
		plot.background = element_rect(fill = "transparent", colour = NA),
		panel.grid.major.y = element_blank(),
		panel.grid.minor.x = element_blank()
	)

>>>>

# Targets from MS workflow
tar_load(
	c(
		accepted_species,
		sanger_sampling,
		coverage_by_rank,
		total_coverage,
		bb_coverage_by_rank,
		bb_total_coverage,
		gb_species_by_year,
		ref_files,
		other_dates,
		crown_ages,
		family_stem_ages
	),
	store = here::here("_targets")
)
```

@@

```{r sampling-stats, message = FALSE}
# Format FTOL sampling as list
tax_coverage <-
	coverage_by_rank %>%
	filter(rank == "major_clade") %>%
	transmute(
		# Be sure to change to factor so list is correct order
		# when splitting
		clade = fct_reorder(name, coverage),
		coverage = percent(coverage, accuracy = 0.1)) %>%
	split(.$clade)

# Format overall FTOL sampling as list
total_sampling <- split(total_coverage, total_coverage$rank)

# Format overall plastome (backbone) sampling as list
bb_total_sampling <- split(bb_total_coverage, bb_total_coverage$rank)
```

@@

plastome_gleich_genera_n <-
	bb_coverage_by_rank %>%
	left_join(select(ppgi_taxonomy, name = genus, order), by = "name") %>%
	filter(order == "Gleicheniales") %>%
	pull(name) %>%
	n_distinct()
```

@@

clade_cols_tib <-
	tibble(
		color = clade_cols,
		major_clade = names(clade_cols)
	) %>%
	mutate(
		code = substr(major_clade, 1, 2),
		code = case_when(
			major_clade == "Saccolomatineae" ~ "Sac",
			major_clade == "Salviniales" ~ "Sal",
			TRUE ~ code
		)) %>%
	assert(is_uniq, code)
```

@@

# Format coverage at order-ish level for plotting
b <-
	coverage_by_rank %>%
	filter(rank == "major_clade") %>%
	left_join(major_clade_exemp_taxa, by = c(name = "major_clade")) %>%
	arrange(phy_order) %>%
	rename(clade = name) %>%
	mutate(
		clade = fct_reorder(clade, phy_order) %>%
			fct_rev(),
		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
	) %>%
	ggplot(aes(x = coverage, y = clade, fill = clade)) +
	geom_col(color = "grey10", size = 0.2) +
	geom_text(
		aes(label = coverage_text),
		hjust = 0,
		nudge_x = 0.01,
		size = 6*.36 # 6 pt font
	) +
	scale_x_continuous(
		labels = percent,
		limits = c(0,1),
		expand = expansion(mult = c(0, 0))
	) +
	scale_fill_manual(values = clade_cols) +
	scale_y_discrete(labels = clade_labs) +
	labs(x = "Coverage") +
	theme_bw() +
	theme(
		axis.text.y = element_markdown(size = 6),
		axis.title.y = element_blank(),
		legend.position = "none",
		plot.background = element_rect(fill = "transparent", colour = NA),
		panel.grid.major.y = element_blank(),
		panel.grid.minor.x = element_blank()
	)
